---
title: 'Prediction of Daily Living Activities using Qualitative Activity Recognition data'
author: "Jay H."
date: "Friday, March 04, 2016"
output: html_document
---

###Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website [here](http://groupware.les.inf.puc-rio.br/har#wle_paper_section#ixzz423nBFGdL).

###Data

The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

The classe variable contains 5 different ways barbell lifts were performed correctly and incorrectly:

- Class A: exactly according to the specification
- Class B: throwing the elbows to the front
- Class C: lifting the dumbbell only halfway
- Class D: lowering the dumbbell only halfway
- Class E: throwing the hips to the front

Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes.

Read more: 

###Objective

The goal of this project is to predict the manner in which people performed barbell lifts. This is the classe variable in the training set. The objective is to create a report describing the prediction models built, use of cross validations techniques and expected sample error for each model. The report also describes the reason for model choices. The report also covers the use your prediction model to predict 20 different test cases.

###Sourcing the data

Packages used for analysis.  
```{r loadlib, message=FALSE, warning=FALSE}
library(caret)
library(randomForest)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
library(dplyr)
```

Load the data into R
```{r loaddata1, cache = TRUE}
# The location where the training data is to be downloaded from
train_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
# The location where the testing data is to be downloaded from
test_url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
# Replace missing values & excel division error strings #DIV/0! #with #'NA'
train_set <- read.csv(train_url, na.strings=c("NA","#DIV/0!",""), header=TRUE)
# Replace missing values & excel division error strings #DIV/0! with 'NA'
test_set <- read.csv(test_url, na.strings=c("NA","#DIV/0!",""), header=TRUE)
# Training data classe variable
summary(train_set$classe)
```

##Partitioning the data for Cross-validation
The training data is split into two data sets, one for training the model and one for testing the performance of our model. The data is partitioned by the classe variable, which is the varible we will be predicting. The data is split into 70% for training and 30% for testing.
```{r partdata}
inTrain <- createDataPartition(y=train_set$classe, p = 0.70, list=FALSE)
training <- train_set[inTrain,]
testing <- train_set[-inTrain,]
dim(training); dim(testing)
```

##Data Processing
Drop the first 7 variables because these are made up of metadata such as time stamp, user which is not relevant for the prediction model.
```{r dataprop}
training <- training[,-c(1:7)]
#Remove NearZeroVariance variables
nzv <- nearZeroVar(training, saveMetrics=TRUE)
training <- training[, nzv$nzv==FALSE]                     
```

Remove variables with no data or 'NA' values. 
Assumption: Drop variables that have 70% or more of the values as 'NA'.
```{r dropna1}
training_nan <- training
for(i in 1:length(training)) {
  if( sum( is.na( training[, i] ) ) /nrow(training) >= .7) {
    for(j in 1:length(training_nan)) {
      if( length( grep(names(training[i]), names(training_nan)[j]) ) == 1)  {
        training_nan <- training_nan[ , -j]
      }   
    } 
  }
}
# Set the new cleaned up dataset back to the old dataset name
training <- training_nan
```
Transform the test_set dataset
```{r getcolname}
# Get the column names in the training dataset
columns <- colnames(training)
# Drop the class variable
columns2 <- colnames(training[, -53])
# Subset the test data on the variables that are in the training data set
test_set <- test_set[columns2]
dim(test_set)
```

###Cross-Validation: Prediction with Random Forest

A Random Forest model is built on the training set. Then the results are evaluated on the test set
```{r cv1, cache=TRUE}
set.seed(12345)
modFit1 <- randomForest(classe ~ ., data=training)
prediction <- predict(modFit1, testing)
cm <- confusionMatrix(prediction, testing$classe)
print(cm)
overall.accuracy <- round(cm$overall['Accuracy'] * 100, 2)
sam.err <- round(1 - cm$overall['Accuracy'],5)
```

The model is `r overall.accuracy`% accurate on the testing data partitioned from the training data. The expected out of sample error is roughly `r sam.err`%.

```{r plot1}
plot(modFit1)
```

In the above figure, error rates of the model are plotted over 500 trees. The error rate is less than `r sam.err`% for all 5 classe.

###Cross-Validation: Prediction with a Decision Tree

```{r cv2}
set.seed(12345)
modFit2 <- rpart(classe ~ ., data=training, method="class")
prediction2 <- predict(modFit2, testing, type="class")
cm2 <- confusionMatrix(prediction2, testing$classe)
print(cm2)
overall.accuracy2 <- round(cm2$overall['Accuracy'] * 100, 2)
sam.err2 <- round(1 - cm2$overall['Accuracy'],2)
```
The model is `r overall.accuracy2`% accurate on the testing data partitioned from the training data. The expected out of sample error is roughly `r sam.err2`%.

Plot the decision tree model
```{r plot2}
fancyRpartPlot(modFit2)
```
###Prediction on the Test Data

The Random Forest model gave an accuracy of `r overall.accuracy`%, which is much higher than the `r overall.accuracy2`% accuracy from the Decision Tree. So we will use the Random Forest model to make the predictions on the test data to predict the way 20 participates performed the exercise.
```{r predict1}
final_prediction <- predict(modFit1, test_set, type="class")
print(final_prediction)
```

###Conclusion

There are many different machine learning algorithm. The Random Forest and Decision Tree models were compared for analysis. For this data, the Random Forest proved to be a more accurate way to predict the manner in which the exercise was done.

###References:
*1. Weight Lifting Exercise DatasetVelloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.*